---
title: "Import and representation of CyCIF tissue microarray data"
author: "Ludwig Geistlinger and Robert Gentleman"
affiliation: Center for Computational Biomedicine, Harvard Medical School
output:
  BiocStyle::html_document:
    self_contained: yes 
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
vignette: >
  % \VignetteIndexEntry{CyCIF TMA}
  % \VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    crop = NULL ## Related to https://stat.ethz.ch/pipermail/bioc-devel/2020-April/016656.html
)
```

# Setup

```{r libs, message = FALSE}
library(cytomapper)
library(vroom)
library(CytoImageList)
library(SpatialExperiment)
```

Start with downloading from and unzipping the dataset from
[synapse](https://www.synapse.org/#!Synapse:syn22345748/files/):

```{r dirs}
data.dir <- "TMA11"
img.dir <- file.path(data.dir, "dearray")
quant.dir <- file.path(data.dir, "quantification")
seg.dir <- file.path(data.dir, "segmentation")
```

# Data

## Metadata

The tissue microarray (TMA) contains cores from 34 cancer, non-neoplastic diseases,
and normal tissues collected from clinical discards under an IRB-supervised protocol.

The TMA was imaged using the cyclic immunofluorescence (CyCIF) method described
in Lin et al 2018. Data were collected with a 20X magnification, 0.75 NA objective 
with 2x2-pixel binning using two multiplex antibody panels (on section 11 "TMA11"
and section 22 "TMA22").

### Markers

A metadata table mapping channel indices to antibodies used for staining, along 
with their QC information.

```{r markers}
marker.file <- file.path(data.dir, "markers.csv")
markers <- read.csv(marker.file)
dim(markers)
head(markers)
```

There are two duplicate markers:

```{r dup-markers}
which(table(markers$marker_name) > 1)
subset(markers, marker_name %in% c("ECAD", "KI67"))
```

Here, we resolve the duplicated marker IDs by attaching an integer suffix. 

```{r resolve-dup-markers}
for(n in c("ECAD", "KI67"))
{
    ind <- markers$marker_name == n
    markers$marker_name[ind] <- paste(markers$marker_name[ind],
                                      1:2,
                                      sep = "_")
}
rownames(markers) <- markers$marker_name
ind <- setdiff(colnames(markers), "marker_name")
markers <- markers[,ind]
```

### Tissue metadata  

For the EMIT dataset, human tissue specimens (from 42 patients) were used to
construct a multitissue microarray (HTMA427). 

Two 1.5-mm-diameter cores were acquired from each of 60 tissue regions with the
goal of acquiring one or two examples of as many tumors as possible (with matched
normal tissue from the same resection when that was feasible), as well as several
non-neoplastic medical diseases involving acute inflammation (for example
diverticulitis and appendicitis), and secondary lymphoid tissues such as tonsil,
spleen and lymph nodes.

Overall, the TMA contained 120 cores plus 3 additional "marker cores", which are
cores added to the TMA in a manner that makes it possible to orient the TMA in images.

A metadata table mapping dearrayed indices (`dearray/*.tif`) to the original
positions in the TMA and the corresponding donor IDs and tissue types.

```{r tissues, message = FALSE}
tissue.file <- file.path(data.dir, "tissues.csv")
tissues <- read.csv(tissue.file)
dim(tissues)
head(tissues)
```

```{r tissue-order}
ind <- order(as.integer(tissues$dearray_id))
tissues <- tissues[ind,]
rownames(tissues) <- NULL
```

Would be good to get some more metadata here on the tissue samples (such
as information on age, sex, and ethnicity of donors, sample location, ...).

```{r tissue-type}
table(tissues$tissue_type)
```

Some information on how many samples of each donor were obtained

```{r donors}
donor.id <- substring(tissues$donor_block_id, 1, 4)
table(donor.id)
```

## Expression data

Spatial feature tables capturing mean marker expression in each cell and basic
morphological features produced by `regionprops()` from
[scikit-image](https://scikit-image.org/docs/0.12.x/api/skimage.measure.html#regionprops)
for each of the 123 cores:

```{r exprs-files, message = FALSE}
files <- list.files(quant.dir, full.names = TRUE)
head(files)
length(files)
```

We read in the data and reshape it to an overall data frame:

```{r exprs-cont}
cont <- lapply(files, vroom::vroom, show_col_types = FALSE)
nr.cells <- vapply(cont, nrow, numeric(1)) 
cont <- do.call(rbind, cont)
cont <- data.frame(cont)
```

```{r exprs-reshape}
core.vec <- basename(files)
core.vec <- sub("^unmicst-", "", core.vec)
core.vec <- sub("\\.csv$", "", core.vec)
core.vec <- as.integer(core.vec)
core.vec <- rep(core.vec, nr.cells)
cont <- cbind(CoreID = core.vec, cont)
```

```{r exprs-dim}
dim(cont)
head(cont)
```

Ok, let's separate this into expression data and cell metadata:

```{r exprs-sep}
ind <- grep("_cellMask$", colnames(cont), value = TRUE)
exprs <- t(cont[,ind])
rownames(exprs) <- sub("_cellMask$", "", rownames(exprs))
stopifnot(rownames(exprs) %in% rownames(markers))
```

```{r coldata}
ind <- setdiff(colnames(cont), ind)
cdat <- cont[,ind]
head(cdat)
```

`X_centroid` and `column_centroid`, and `Y_centroid` and `row_centroid` seem to be
identical, so let's drop one of those:

```{r spatial-coords}
all(cdat$X_centroid == cdat$column_centroid)
all(cdat$Y_centroid == cdat$row_centroid)
ind <- setdiff(colnames(cdat), c("column_centroid", "row_centroid"))
cdat <- cdat[,ind]
```

Obtain spatial coordinates mat:

```{r spatial-coords-mat}
xy <- c("X_centroid", "Y_centroid")
smat <- as.matrix(cdat[,xy])
colnames(smat) <- c("x", "y")
ind <- setdiff(colnames(cdat), xy)
cdat <- cdat[,ind]
head(cdat)
```

```{r tissue-cdat-sync}
colnames(tissues)[1] <- colnames(cdat)[1] <- "sample_id"
ind <- match(cdat[,"sample_id"], tissues[,"sample_id"])
head(cdat)
```

Create `SpatialExperiment` from individual components:

```{r create-spe}
spe <- SpatialExperiment::SpatialExperiment(assays = list(exprs = exprs),
                                            colData = cdat,
                                            rowData = markers,      
                                            spatialCoords = smat)
```

Where to we store sample-level metadata in a `SingleCellExperiment` / `SpatialExperiment`?
People seem to just replicate sample-level metadata for each cell beloning
to one sample. That doesn't seem to be the most memory-efficient approach though.

```{r tissue-metadata}
metadata(spe)$tissues <- DataFrame(tissues)
spe
```

Inspect the `SpatialExperiment`:

```{r inspect-spe}
assay(spe)[1:5,1:5]
colData(spe)
rowData(spe)
head(spatialCoords(spe))
metadata(spe)$tissues
```

## Image data

### Raw images 

Raw images of 123 individual cores in each TMA image.
These have been stitched and registered from the original TMA images using
[ASHLAR](https://github.com/labsyspharm/ashlar) and dearrayed using 
[Coreograph](https://github.com/HMS-IDAC/UNetCoreograph). 

As each of the 123 images is >1 GB, and will thus be too big to fit into memory
on most machines, we  store them as HDF5 first. This conversion to HDF5 is slow
and is best carried out in parallel on eg a computer cluster.

```{r hdf5-files}
hdf5.dir <- file.path(img.dir, "hdf5")
if(!file.exists(hdf5.dir)) dir.create(hdf5.dir)
```

```{r load-images, eval = FALSE}
img.list <- cytomapper::loadImages("dearray",
                                   pattern = ".tif",
                                   on_disk = TRUE,
                                   h5FilesPath = hdf5.dir)
```

Once we've converted the tif files to h5 files, we can load them chunkwise with
loading data on demand:

```{r load-images-hdf5}
img.list <- cytomapper::loadImages(hdf5.dir)
ind <- order(as.integer(names(img.list)))
img.list <- img.list[ind]
img.list
```

Let's add names to the channels

```{r channel-names}
channelNames(img.list) <- rownames(spe)
```

Let's inspect the a subset of images and a subset of markers for demonstration
purposes:

```{r plot-pixels, fig.width = 10, fig.height = 10}
cytomapper::plotPixels(img.list[1:8],
                       colour_by = channelNames(img.list)[1:6])
```

Which tissues are these:

```{r tissues-plotted}
subset(tissues, sample_id %in% as.character(1:8))
```

### Masks

Segmentation masks of cells and nucleis computed with a deep learning model
based on a [UNet architecture](https://github.com/HMS-IDAC/UnMicst).

This requires some restructuring of the `segmentation` subdirectory of the
primary data release - it contains a subdirectory for each core with files
named `cellMask.tif` and `nucleiMask.tif`, but what we actually need is a
directory for both mask types.

```{r masks}
cell.mask.dir <- file.path(seg.dir, "cell_masks")
nuclei.mask.dir <- file.path(seg.dir, "nuclei_masks")
```

We can now copy the files:

```{r seg-dir-restruct, eval = FALSE}
for(f in cell.mask.files)
{
    spl <- unlist(strsplit(f, "/"))[3]
    i <- sub("unmicst-", "", spl)
    of <- file.path(cell.mask.dir, paste(i, "tif", sep = "."))
    file.copy(f, of)
}
for(f in nuclei.mask.files)
{
    spl <- unlist(strsplit(f, "/"))[3]
    i <- sub("unmicst-", "", spl)
    of <- file.path(nuclei.mask.dir, paste(i, "tif", sep = "."))
    file.copy(f, of)
}
```

Read the masks:

```{r read-masks}
cell.masks <- cytomapper::loadImages(cell.mask.dir, as.is = TRUE)
nuclei.masks <- cytomapper::loadImages(nuclei.mask.dir, as.is = TRUE)
```

Sort in numeric core order:

```{r order-masks}
ind <- order(as.integer(names(cell.masks)))
cell.masks <- cell.masks[ind]
mcols(cell.masks)$sample_id <- names(cell.masks)

ind <- order(as.integer(names(nuclei.masks)))
nuclei.masks <- nuclei.masks[ind]
mcols(nuclei.masks)$sample_id <- names(nuclei.masks)
```

Let's plot some example cores:

```{r plot-cell-masks}
cytomapper::plotCells(cell.masks[c(1, 57, 114)],
                      object = spe,
                      img_id = "sample_id",
                      cell_id = "CellID", 
                      colour_by = channelNames(img.list)[1:6],
                      exprs_values = "exprs")
```

```{r plot-nuclei-masks}
cytomapper::plotCells(nuclei.masks[c(1, 57, 114)],
                      object = spe,
                      img_id = "sample_id",
                      cell_id = "CellID", 
                      colour_by = channelNames(img.list)[1:6],
                      exprs_values = "exprs")
```

Which tissues are these:

```{r tissues-plotted2}
subset(tissues, sample_id %in% as.character(c(1, 57, 114)))
```

We can also plot the whole TMA, ie. all 123 cores:

```{r plot-TMA, fig.width = 12, fig.height = 12}
cytomapper::plotCells(cell.masks,
                      object = spe,
                      img_id = "sample_id",
                      cell_id = "CellID",
                      colour_by = channelNames(img.list)[1:6],
                      exprs_values = "exprs")
```
